Data Specialization for a Staged Lambda Calculus

### 2015-07-07 paper/semantics.tex
adding more structure to splitting section

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
Consequently, this paper does not describe in any detail the second pass of evaluation.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
Thus, $\steppure e {e'}$ searches for redexes and reduces them in the standard way, and a ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

>>>>

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
Consequently, this paper does not describe in any detail the second pass of evaluation.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgement does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code (compatibility rules) 
where it either performs a beta reduction (reduction rules) or marks the term as being evaluated (value marking rules). 
A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

### 2015-07-07 paper/splitting.tex
adding more structure to splitting section

The goal of a stage splitting translation is to send programs in a multistage language
to a form where the stages are seperated at the top level.
To be valid, such a translation must meet two requirements:
at values it must not lose information,
and at terms it must respect the semantics.

For a two-stage language like \lang, this means translating a multistage
term into a pair of monostage terms
which respectively embody the first-stage and second-stage content of the original.
Since \lang\ has three classes of multistage term, 
our splitting algorithm has three forms: 
\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
The rules of all three judgements are given in \ref{fig:termSplit}, 
and they operate locally on the structure of the term.
Note that there's no need to provide forms of splitting for ground terms or residuals,
since those classes of term are already monostage by construction.

Before discussing the validity of these translations, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting relation is a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

For any $\coltwo e A$
which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
splitting $e$ (via $\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.

This output is a multistage program in its own right.
It evaluates by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

>>>>

The goal of a stage splitting translation 
is to a send program $\mathcal P$ in a multistage language
to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
the same answer under their respective semantics.

Since \lang\ has three classes of multistage term, 
our splitting algorithm has three forms: 
$\splittwo e A p l r$ for \bbtwo-terms, 
$\splitone e A c {l.r}$ for \bbonem-terms, and 
$v \vsplito \mval i q$ for partial values.
In each form the output has two parts, corresponding to the
first-stage ($p$, $c$, and $i$) and second-stage ($l.r$, and $q$) content of the input.
Note that there's no need to provide forms of splitting for ground terms or residuals,
since those classes of term are already monostage by construction.

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
They are simply recursive on structure of the term, 
implying that the splitting algorithm runs in linear time.

Before discussing the correctness of these translations, 
we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting judgement defines a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

Top level programs in \lang\ are given by a term $\coltwo e A$.
Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
We write this as:
$
e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q
$.

Splitting $e$ (via $\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.
This output is a multistage program in its own right.
It evaluates by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@
